The Private School Network: Recruiting Visits to Private High Schools by Public and Private Universities

### 2023-11-16 r_r/manuscript_rihe.Rmd
revising discussion

[PUT THIS IN THE DISCUSSION SECTION] Finally When both private and public universities are targeting private schools, competition for these students is likely to increase. This could be an explanation for why many private universities are struggling to meet enrollment goals

student bodies that predonominantly affluent, white student bodies.

@@

Figure \@ref(fig:race-privu-privhs-pubhs) focuses on selective private universities, showing the racial composition of visited private high schools (top panel) to the racial composition of visited public high schools (bottom panel). Results are very similar to those of public universities. Visited private schools enrolled a much lower share of Black/Latinx/Native students than visited public schools. For Emory, Northwestern, Tufts, Tulane, Case Western Reserve, and University of Denver, at least 75% of visited private schools enrolled less than 20% Black/Latinx/Native students.

Several studies suggest that universities -- particularly elite private universities -- pursue racial diversity by recruiting non-white students who attend predominantly white private high schools [e.g., @RN4495]. Additionally, scholarship that engages with enrollment in private schools by Black students suggests that many high-income Black families began sending their children to elite, predominantly white private schools once these schools were willing to enroll Black students [@RN4724]. Therefore, universities may visit predominantly white private high schools in search of non-white students. One concern about this practice is that non-white students at elite private schools may be in high demand by many universities, but similarly achieving non-white students at predominantly non-white public schools might be ignored [THIS PARAGRAPH SHOULD PROBABLY GO TO DISCUSSION].

@@

Research question 2 asks about the characteristics of visited private high schools. We analyzed geographic region, religious affiliation, high school ranking, 12th grade enrollment, and racial composition. Both public and private universities in our sample tended to make a disproportionate number of visits in their home geographic region, with highly ranked universities making more visits to schools in other regions. The majority of universities -- even those located outside the South --  visited a large number of private schools located in the South. Unsurprisingly, sectarian private universities made a large share of visits to private high schools of a similar religious affiliation. Given that public universities are nonsectarian organizations, a surprising finding is that the majority of public universities in our sample made a large share of visits to Catholic or Christian private schools.

@@

Across the universe of private high schools, most schools enroll relatively few students who identify as Black, Latinx, or Native. For the public and private universities in our sample, the set of visited private schools enrolled fewer Black, Latinx, and Native students than the national universe of private schools. By contrast, the public high schools visited by public universities tended to enroll a much higher share of students who identify as Black, Latinx, or Native.

Research question 3 asks about overlap in the private school recruiting networks of public and private universities. Within the sample of public universities, we observed substantial overlap and variation in recruiting networks. For example, the recruiting networks of the University of Arkansas and the University of Kansas share substantial overlap and are very different from the recruiting networks of UMass Amherst and Rutgers, which share substantial overlap with one another. Across the entire sample of public and private universities, we also observed substantial overlap and variation in recruiting networks. Whether we are categorizing universities by cluster analysis or by proximity on the network graph, we observe several different subsets of universities. Each of these subsets is associated with a different approach to private school visits and -- with the exception of a few universities like UC-Riverside and SUNY Stony Brook -- each of these subsets contained both public and private universities.

## Privatization of public research universities

These findings contribute new insights about the privatization of public research universities. A weakness of the scholarly literature on privatization is that few studies compare public and private universities with respect to a behavior that is associated with private universities [@RN4543]. We conceptualize recruiting visits to private schools as a behavior that is consistent with the social reproduction orientation of selective private universities and contrary to the historic social mobility mission associated with public research universities. Although we cannot make statements about the population of public research universities (ours is a convenience sample rather than a random sample), we can describe variation in how the public universities in our sample went about visiting private schools.

At one end of the continuum lie public research universities that do not play the game, for example UC-Riverside and SUNY Stony Brook, visiting private schools at a lower rate than public schools (Figure \@ref(fig:actual-proportional-pubu)). For public universities that visited a substantial number of private schools, we observe substantial variation in the scale of visits and the characteristics of schools they target. While all private universities in our sample visited private schools at a much higher rate than public schools, we also see substantial variation amongst private universities in the scale and characteristics of private school visits. Amongst the network of public and private universities, Figure \@ref(fig:plot-1mode-u) shows that three clusters emerge -- each containing some mix of public and private universities. The universities within a cluster have some overlap in the kinds of schools visited (e.g., TCU and University of South Carolina), but also share substantial overlap with particular universities in different clusters (e.g., Tufts and University of Pittsburgh).

In making large-scale visits to out-of-state private high schools, we argue that public research universities are mimicking the behavior of selective private universities in order to compete for students that have been traditionally targeted by private universities. Our rationale for this statement is that selective private universities have been recruiting heavily from private schools for a long time [@RN4642; @RN1565; @RN4407]. By contrast, the pursuit of out-of-state students is a relatively new phenomenon for most public research universities.

When making decisions about which private schools to visit, public research universities seem to be mimicking the behavior of particular private universities rather than mimicking the behavior of private universities writ large. Several public universities seem to be competing directly for students targeted by a particular private university. For example, UC-Berkeley visited 27.9% of the schools visited by Northwestern, 28.9% of the schools visited by Tufts, and 27.8% of the schools visited by Emory (Figure \@ref(tab:privu-percent-visited-by-pubu-matrix)). We reason UC-Berkeley is competing directly for students because all four universities are in the same broad echelon of prestige. Other potential examples of direct competition include the University of Georgia targeting schools visited by Emory, SMU, TCU, and Tulane, or the University of Arkansas targeting schools visited by Baylor.

@@

Finally, all public universities that visited a substantial number of private schools tended to visit schools that enrolled few Black, Latinx, or Native students. This finding is substantially driven by the fact that the national universe of private schools tends to enroll a much lower share of Black, Latinx, or Native students than the universe of public schools. However, the set of schools visited by the public universities in our sample enrolled fewer Black, Latinx, and Native students than the universe of private schools. Although we do not have data on household income, it is likely that the majority of students are wealthy at the private schools visited by public universities in our sample.

Therefore, while public universities playing the private school game differ in terms of the set of schools they target, they all seem to be targeting schools that enroll mostly affluent students and few students who identify as Black, Latinx, or Native. To the extent that these recruiting efforts yield dividends in terms of incoming freshmen, playing the private school game is likely to change the enrollment composition of public research universities in ways that run contrary to racial and socioeconomic diversity.

>>>>

@@

Figure \@ref(fig:race-privu-privhs-pubhs) focuses on selective private universities, showing the racial composition of visited private high schools (top panel) to the racial composition of visited public high schools (bottom panel). Results are very similar to those of public universities. Visited private schools enrolled a much lower share of Black/Latinx/Native students than visited public schools. For Emory, Northwestern, Tufts, Tulane, Case Western Reserve, and University of Denver, at least 75% of visited private schools enrolled less than 20% Black/Latinx/Native students

@@

Research question 2 asks about overlap in the private school recruiting networks of public and private universities. We observed substantial overlap and variation in recruiting networks of public universities. For example, the recruiting networks of the University of Arkansas and the University of Kansas share substantial overlap and are very different from the recruiting networks of UMass Amherst and Rutgers, which share substantial overlap with one another. Amongst the network of public and private universities, we observe substantial overlap and variation in recruiting networks. Figure \@ref(fig:plot-1mode-u) shows that three clusters emerge -- each containing some mix of public and private universities. The universities within a cluster have some overlap in the kinds of schools visited (e.g., TCU and University of South Carolina), but also share substantial overlap with particular universities in different clusters (e.g., Tufts and University of Pittsburgh). Whether we are categorizing universities by cluster analysis or by proximity on the network graph, we observe several different subsets of universities. Each of these subsets is associated with a different approach to private school visits and -- with the exception of a few universities like UC-Riverside and SUNY Stony Brook that don't play the private school game -- each of these subsets contained both public and private universities.

Research question 3 asks about the characteristics of visited private high schools. We analyzed geographic region, religious affiliation, high school ranking, 12th grade enrollment, and racial composition. Both public and private universities in our sample tended to make a disproportionate number of visits in their home geographic region, with highly ranked universities making more visits to schools in other regions. The majority of universities -- even those located outside the South --  visited a large number of private schools located in the South. Unsurprisingly, sectarian private universities made a large share of visits to private high schools of a similar religious affiliation. Given that public universities are nonsectarian organizations, a surprising finding is that the majority of public universities in our sample made a large share of visits to Catholic or Christian private schools.

@@

Across the universe of private high schools, most schools enroll relatively few students who identify as Black, Latinx, or Native. For the public and private universities in our sample, the set of visited private schools enrolled fewer Black, Latinx, and Native students than the national universe of private schools. By contrast, the public high schools visited by public universities tended to enroll a much higher share of students who identify as Black, Latinx, or Native.n [ADD A SENTENCE OR TWO HERE]

## Implications for Scholarship

This study engages with two, previously disconnected literatures. Scholarship on nonresident enrollment observes that many public research universities began growing nonresident enrollment following declines in state appropriations []. Given that nonresident tuition price is often two- to three-times resident price, it is unsurprising that nonresident students tend to be more affluent than resident students, are less likely to be first-generation students, and are less likely to be Black or Latinx [CITE]. Public research universities recruit nonresident students using institutional financial aid packages [CITE] and through recruiting interventions that target earlier stages in the enrollment funnel. Using nearly the same sample of 15 public research universities used in this study, Author (XXXX) found that 12 universities made more out-of-state visits than in-state visits and visits to out-of-state public high schools focused on affluent, predominantly white schools. Public research universities have a financial incentive to target private school students, who tend to come from affluent households able to pay nonresident tuition price without need-based aid. This study extends the literature on nonresident students by analyzing the scale of visits to private schools, the overlap and divergence in recruiting networks between public and private universities, and the characteristics of visited private schools.

We also engage with scholarship on linkages between private schools and private universities, a linkage that sociologists have observed for many decades. Private school students comprise a disproportionate share of enrollment at selective private universities [CITE; AND AUTHOR CALCULATIONS]. Practically, these linkages exist because private schools want to send their students to "good" universities while selective private universities want to enroll students who can pay private university tuition and who have desired academic profile and/or extracurricular activities. However, prior research tends to be focused on elite schools and universities [CITE KAHN; CHETTY; CHARTER+BARTER] and does not systematically analyzed connections between private schools and private universities that are selective, but not elite. This study extends the literature by investigating private school visits for a sample of 14 selective private universities that exhibit variation in geographic location, religious affiliation, and ranking.

Our more important scholarly contribution is showing that these two separate stories are both parts of a broader narrative about the changing nature of competition for students from privileged schools and communities. Selective private universities have been recruiting  heavily from private schools for a long time [@RN4642; @RN3519; @RN4640]. By contrast, the pursuit of out-of-state private school students by public research universities is a relatively new phenomenon and represents competition in the traditional hunting grounds of selective private universities.

Our analyses shed light about the nature of this competition. When making decisions about which private schools to visit, public research universities seem to be mimicking the behavior of particular private universities rather than mimicking the behavior of private universities writ large. Several public universities seem to be competing directly for students targeted by a particular private university. For example, UC-Berkeley visited 27.9% of the schools visited by Northwestern, 28.9% of the schools visited by Tufts, and 27.8% of the schools visited by Emory (Figure \@ref(tab:privu-percent-visited-by-pubu-matrix)). We reason UC-Berkeley is competing directly for students because all four universities are in the same broad echelon of prestige. Other potential examples of direct competition include the University of Georgia targeting schools visited by Emory, SMU, TCU, and Tulane, or the University of Arkansas targeting schools visited by Baylor.

@@

While public universities playing the private school game differ in terms of the set of schools they target, they all seem to be targeting schools that enroll mostly affluent students and few students who identify as Black, Latinx, or Native. To the extent that these recruiting efforts yield dividends in terms of incoming freshmen, playing the private school game is likely to change the enrollment composition of public research universities in ways that run contrary to racial and socioeconomic diversity.

## Implications for Practice

Market research suggests that off-campus recruiting visits will remain a vital means of connecting to students and guidance counselors, just as they were prior to the Covid pandemic. @RN4664 found that "travel" was the top-ranked top-ranked marketing/recruiting expenditure item for both public universities and private universities. During the height of Covid, off-campus recruiting visits ceased and were partially replaced by virtual counselor events [CITE]. Admissions counselors are now traveling again. @RN4896 found that travel was the fifth ranked expenditure item for public universities and the second ranked expenditure item for private universities. When asked about areas of future expenditure growth, travel was the top ranked item for public universities and the second ranked item for private universities @RN4896, suggesting that universities are increasing travel following the decline of the pandemic.

Additionally, the EAB (XXXX) survey of high school counselors asked "how do you prefer to receive information from colleges?" High school visits by college representatives was the highest rated "channel" for both public school counselors and (especially) for private school counselors. By contrast, virtual visits by admissions representatives received low ratings by high school guidance counselors. Finally, @RN4895 asked universities to rank the most effective "outreach strategies for high school counselors." "Calling after visits" was deemed effective for 97% of private universities (the top strategy) and was deemed effective by 88% of public universities. What insights do we glean from these market research findings? In the era of mass marketing and data science, the pursuit of traditional college students from privileged schools and communities is still a hand-to-hand business and off-campus recruiting visits remain a vital means of recruiting these students.

Considering the practical importance of off-campus recruiting visits, what are the practical implications of our research? Prior scholarly and market research [] shows that students are sensitive to recruiting interventions, including recruiting visits. This means that racial and socioeconomic composition of the schools that a university chooses to visit affects the racial and socioeconomic composition of the university. Our research on visits to public high schools [CITE] and this study on visits to private schools finds that our sample of public research universities and selective private universities tend to focus their recruiting efforts on affluent and predominantly white schools and communities.    Some universities may publicly proclaim a commitment to social mobility and privately pursue an intentional recruiting strategy that targets affluent schools and communities. We offer practical guidance for universities that may be unintentionally implementing a recruiting strategy tied to social reproduction.

Several studies suggest that universities -- particularly elite private universities -- pursue racial diversity by recruiting non-white students who attend predominantly white private high schools [e.g., @RN4495]. Additionally, scholarship that engages with enrollment in private schools by Black students suggests that many high-income Black families began sending their children to elite, predominantly white private schools once these schools were willing to enroll Black students [@RN4724]. Therefore, universities may visit predominantly white private high schools in search of non-white students. One concern about this practice is that non-white students at elite private schools may be in high demand by many universities, but similarly achieving non-white students at predominantly non-white public schools might be ignored

In the aftermath of the Supreme Court decision on race conscious admissions, @RN4954 published a report titled *Strategies for Increasing Diversity and Opportunity in Higher Education.*  The report recommended that universities diversify the set of high schools they recruit from:

>> Rather than recruiting primarily from private and public high schools composed of predominantly high-income students, institutions can expand and prioritize their outreach to high schools with substantial populations of low-income students and students of color. As part of expanding outreach, institutions can assign admissions recruiters to high schools from which they have not typically recruited to provide equal opportunity for all students

We suspect that many universities may have incomplete knowledge about which schools they recruit from and how these schools compare to those they do not recruit from. Our rationale for this statement is that employee turnover in the enrollment management profession is quite high [CITE HOOVER] and many universities rely on expertise from third-party enrollment management consultancies [AUTHOR]. Consistent with this idea, when issued public records requests to collect data on "student list" purchases from about 100 public universities [CITE], we found that many public universities did not know who or how they purchased student lists, typically because the person who did this had left the university and/or the job was outsourced to a consultancy.

For university leaders concerned about racial and socioeconomic equity in college access, we recommend conducting an internal audit of recruiting practices. How do the high schools and community colleges they visit differ racially, socioeconomically, and geographically from those they do not visit? How do the student list prospect profiles they purchase differ racially, socioeconomically, and geographically from those they do not purchase? We also recommend that students, faculty, and university stakeholders ask this question of the university president. It may be that the president is unaware and must ask the enrollment management office, which may begin a useful conversation. 
While universities may be aware of the composition of the schools they visit and the prospect profiles they purchase, we suspect that they may be less aware of how the composition of targeted schools/communities/students compares to the composition of not targeted schools/communities/students. If internal equity audits point to areas of inequity in recruiting behavior, university stakeholders can begin a discussion of how to address this issue.
